---
title: "The Effect of Education Level on Income Inequality"
author: "Names + Studentnumbers"
date: "`r Sys.Date()`"
output: pdf_document
---

#Set up your environment
```{r}
setwd("~/")
```

```{r packages_install, include = FALSE}
install.packages("tidyverse")
install.packages("renv")
install.packages("dplyr")
```

```{r}
library(tidyverse)
library(renv)
library(dplyr)
```

```{r}
packageVersion("tidyverse")
packageVersion("renv")
packageVersion("dplyr")
```

#Title Page
Carlijn Calori, Leah Delikát, Fadhil Dhafir, Sten Groen, Julia Koeleman, Anne Schrama, Marie-Louise Stevens, Sophia Zentgraf
Tutorial: Group 1
J.F. Fitzgerald

#Part 1 - Identification of the Social Problem

##1.1 Describe the Social Problem
Income inequality in the USA is a social problem because it limits equal opportunities, increases social tensions, and harms health and democracy.The Pew Research Center warns that the middle class is shrinking rapidly, while the rich grow richer.In addition, The OECD ranks the U.S. among the most unequal developed nations. These two studies emphasize the importance of this problem ABCDEF. 


https://www.pewresearch.org/social-trends/2020/01/09/trends-in-income-and-wealth-inequality/?
https://www.oecd.org/en/publications/inequality-and-poverty-in-the-united-states_5k46957cwv8q-en.html

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading data}
library(readxl)
Distribution <- read_excel("data/Frank_WID_2020.xls", sheet = 3)
Education <- read_excel("data/tabn012.xls", col_types = c("text", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip","skip", "skip", "skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip"))

```

## 3.1 Summary of the datasets

## 3.2 Describing the type of variables in the datasets

This is our merged data set. In the first column you can see what year the data refers to. Next to it is the column with the State_ID. We put the states in alphabetical order and then started numbering from 1 to 100 (example: Alabama is numbered 1 in 1990 and 51 (1+50) in 2000). 

The next 5 columns show the percentages of total income that a group earns. For example, the first column shows that the top 10 percent of big earners, earn 39.9% of total income, from the state. This is the same for the 4 other columns, but here the groups get smaller and smaller. 

Columns 9 through 14 show how much you have to earn to belong to that percentage of big earners. For example, in 1990 you had to earn more than $101,762.24 to be among the top 10 big earners in Alabama. Did you want to be among the top 5 in 1990? Then you had to earn more than $131,506.30. 

The last two columns show level of education. These variables are in percent. So for example, in 1990, 66.9% of the population in Alabama had a high school diploma or higher and 15.7% had a bachelor's degree or higher. 



# Part 3: Quantifying 

## 3.1 Data Cleaning

Only the years 1990 to 2000 are kept in the Distribution dataset. This is because the Education dataset includes data only for 1990 and 2000. TO ensure the datasets can be merged and compared properly, just 1990, 2000, and the years in between are included. 

In addition, two rows are removed, as they are not a U.S. state. One of the rows corresponds to the entire United States, while the other represents the District of Columbia.
```{r}
Distribution <- Distribution %>%
  filter(Year %in% 1990:2000, 
         State != c("United States", "District of Columbia"))
```

For the Education dataset, unnecessary rows are removed. These rows contain no data and consist only of blank spaces between entries. Also, two rows representing the entire United States and the District of Columbia are removed. 

To simplify the datatset, columns are renamed to reflect the type of data they contain, and any dots following state names are removed.

For the final step of cleaning, before merging the two datasets, the Education dataset needs to be converted to a long format. Currently, it contains four separate variables for the years 1990 and 2000, but these need to be combined into a single column with the years listed vertically. This transformation allows the dataset to be merged properly with the Distribution dataset, which is already in long format.

```{r}
Education <- Education[ -c(1:13, 14, 20, 24, 26, 32, 38, 44, 50, 56, 62, 68, 75:79), ]
``` 

```{r}
names(Education)[1:5] <- c("State", "Highschool1990", "Highschool2000", "Bachelor1990", "Bachelor2000")
```

```{r}
Education <- Education %>% 
  mutate(State = State %>%
           str_replace_all("…+", "") %>%
           str_replace_all("\\.+", "") %>%
           str_trim()) # to delete all the dots after the state 
```

```{r}
df1990 <- data.frame(State = Education$State, Year = 1990, Highschool = Education$Highschool1990, Bachelor = Education$Bachelor1990, stringsAsFactors = FALSE)

df2000 <- data.frame(State = Education$State,Year = 2000,Highschool = Education$Highschool2000,Bachelor = Education$Bachelor2000,stringsAsFactors = FALSE)

Education <- rbind(df1990, df2000)

Education <- Education %>%
  mutate(across(3:4, ~ as.numeric(.)))

```

```{r}
Allmerged <- Distribution %>%
  full_join(Education, by = c("State", "Year"))

Allmerged <- Allmerged %>%
  select(-st) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))
```

## 3.2 Creating new variables
To measure inequality: top 10% divided by bottom 90%.Not an ideal measuring, but we don't have more data on the distribution if income.

```{r}
Allmerged <- Allmerged %>%
  mutate(Top10_vs_bottom90 = round(Top10_adj/(100-Top10_adj), 2))
```

##3.2 Spatial variation of change in income inequality
```{r}
# Make sure these packages are installed and loaded
library(dplyr)
library(tidyr)
library(maps)
library(ggplot2)

diff_data <- Allmerged %>%
  filter(Year %in% c(1990, 2000)) %>%
  select(State, Year, Top10_vs_bottom90) %>%
  pivot_wider(names_from = Year, values_from = Top10_vs_bottom90) %>%
  mutate(
    # Extract only the real state name before any dots or extra characters
    state_lower = tolower(gsub("^([a-zA-Z ]+).*", "\\1", State)),
    state_lower = trimws(state_lower),
    diff = `2000` - `1990`
  ) %>%
  filter(state_lower != "district of columbia")

us_states <- map_data("state")
map_plot_data <- left_join(us_states, diff_data, by = c("region" = "state_lower"))

# Check: should now match!
print(setdiff(sort(unique(us_states$region)), sort(unique(diff_data$state_lower))))
print(setdiff(sort(unique(diff_data$state_lower)), sort(unique(us_states$region))))

ggplot(map_plot_data, aes(long, lat, group = group, fill = diff)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red", midpoint = 0,
    name = "Difference\n(2000 - 1990)"
  ) +
  labs(
    title = "Change in Income Inequality by State",
    subtitle = "Top 10% vs Bottom 90% (2000 - 1990)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```



##3.3 Temporal visualization of income inequality
```{r}
df_summary <- Allmerged %>%
  group_by(Year) %>%
  summarise(median_ineq = median(Top10_vs_bottom90),
            mean_ineq = mean(Top10_vs_bottom90))

ggplot(df_summary, aes(x = Year)) +
  geom_line(aes(y = median_ineq), color = "blue", 
            linewidth = 1.2, 
            lineend = "round") +
  geom_line(aes(y = mean_ineq), 
            color = "red",
            linewidth = 1.2,
            lineend = "round") +
  labs(title = "Income Inequality per year", 
       subtitle = "Median (blue) vs Mean (red)",
       x = "\nYear", 
       y = "Top 10% / Bottom 90% ratio\n") +
  scale_x_continuous(limits = c(1990, 2000), breaks = 1990:2000) +
  scale_y_continuous(breaks = seq(0.5, 0.8, by = 0.02)) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5))


```

##3.4 To visualize income inequality across different percentages of bachelor's degree, we divide them into quartiles.
```{r}
library(ggplot2)

#Create quartile-based education groups per year
Allmerged <- Allmerged %>%
  group_by(Year) %>%
  mutate(Quartile = ntile(Bachelor, 4),
         Bachelor_level = case_when(
           Quartile == 1 ~ "Low",
           Quartile == 2 ~ "Medium Low",
           Quartile == 3 ~ "Medium High",
           Quartile == 4 ~ "High"
         ))
Allmerged <- Allmerged %>%
  mutate(Quartile = NULL)

# Make sure Education is a factor with the right order
Allmerged$Bachelor_level <- factor(
  Allmerged$Bachelor_level,
  levels = c("Low", "Medium Low", "Medium High", "High")
)

#We can use indexing or create new dataframes for 1990 and 2000 to create boxplots per year
library(ggplot2)

ggplot(data = Allmerged[Allmerged$Year == 1990, ], aes(x = Bachelor_level, y = Top10_vs_bottom90)) +
  geom_boxplot() +
  labs(
    x = "\nEducation Group (Based on Bachelor % Quartiles)",
    y = "Income Inequality (Top 10% vs Bottom 90%)\n",
    title = "Effect of Education Level on Income Inequality"
  )

```

##3.5 Event analysis
```{r}
df_summary <- Allmerged %>%
  group_by(Year) %>%
  summarise(median_ineq = median(Top10_vs_bottom90),
            mean_ineq = mean(Top10_vs_bottom90))

ggplot(df_summary, aes(x = Year)) +
  geom_line(aes(y = median_ineq), color = "blue", 
            linewidth = 1.2, 
            lineend = "round") +
  geom_line(aes(y = mean_ineq), 
            color = "red",
            linewidth = 1.2,
            lineend = "round") +
  labs(title = "Income Inequality per year", 
       subtitle = "Median (blue) vs Mean (red)",
       x = "\nYear", 
       y = "Top 10% / Bottom 90% ratio\n") +
  scale_x_continuous(limits = c(1990, 2000), breaks = 1990:2000) +
  scale_y_continuous(breaks = seq(0.5, 0.8, by = 0.02)) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)) +
  geom_vline(xintercept = 1994) +
  annotate("text", x = 1995, y = 0.76, size = 3, label = "Introduction\nof NAFTA")

```
The North American Free Trade Agreement (NAFTA) between the U.S., Mexico, and Canada came into effect in January 1994 with the goal of creating a trilateral trade bloc to increase economic integration, boost trade, and stimulate growth by removing trade barriers and import tariffs. As a result, exports and imports became cheaper, and a significant share of U.S. manufacturing was relocated to Mexico due to lower labor costs (SOURCE).

NAFTA established the world’s largest free trade zone, but not everyone benefited equally. Higher-income groups in the U.S. gained the most, as large corporations experienced increased profits that mainly flowed to shareholders and high-income individuals (This trend explains the rise in the mean income, while the median grew more slowly)(SOURCE).

While the NAFTA boosted overall GDP, it disproportionately affected lower-income groups. The United States lost numerous manufacturing jobs (Estimated at 682,900 by 2010) as companies relocated production to Mexico to cut labor costs (Scott, 2011). This acceleration of outsourcing significantly impacted American manufacturing and low-skilled workers. Consequently, some workers saw their income stagnate or decline, whereas individuals in management and high-skill sectors experienced income growth (SOURCE). As a result, income inequality widened, further expanding the gap between the bottom 90 percent and the top earners.

However, it is important to recognize that the rise in income inequality cannot be attributed solely to NAFTA. Factors such as technological advancements, shifts in labor market policies and broader globalization have also played significant roles in driving this trend.

#4 Discussion

##4.1 Discuss our findings


#5 Reproductibility

##5.2 Reference list
Pew Research Center. (2020, January 9). Trends in U.S. income and wealth inequality. Pew Research Center.
Retrieved from https://www.pewresearch.org/social-trends/2020/01/09/trends-in-income-and-wealth-inequality/
  
  Denk, O., Hagemann, R., Lenain, P., & Somma, V. (2013, May 27). Inequality and poverty in the United States: Public policies for inclusive growth (OECD Working Paper No. 1052). OECD.
Retrieved from https://www.oecd.org/en/publications/inequality-and-poverty-in-the-united-states_5k46957cwv8q-en.html

Scott, R. E. (2011, 3 mei). Heading South: U.S.-Mexico trade and job displacement after NAFTA (Briefing Paper #308). Economic Policy Institute. http://www.epi.org/publications/entry/briefing_paper_308/





