---
title: "The Effect of Education Level on Income Inequality"
author: "Carlijn Calori (2860228), Leah Delikát (2813155), Fadhil Dhafir (2847129), Sten Groen (2813781), Julia Koeleman, Anne Schrama (2812834), Marie-Louise Stevens (2814083), Sophia Zentgraf"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r include=FALSE}
setwd("~/")
```

```{r packages_install, include=FALSE}
install.packages("tidyverse")
install.packages("renv")
install.packages("dplyr")
install.packages("readxl")
```

```{r include=FALSE}
library(tidyverse)
library(renv)
library(dplyr)
library(readxl)
```

```{r include=FALSE}
packageVersion("tidyverse")
packageVersion("renv")
packageVersion("dplyr")
packageVersion("readxl")
```

# Title Page

Carlijn Calori, Leah Delikát, Fadhil Dhafir, Sten Groen, Julia Koeleman, Anne Schrama, Marie-Louise Stevens & Sophia Zentgraf

Tutorial: Group 1

Tutor: J.F. Fitzgerald

# 1 Identification of the Social Problem

## 1.1 Describe the Social Problem

Income inequality is a serious social issue in the United States. Over the past few decades, the gap between the highest and lowest earners has grown a lot, leading to problems on social, economic, and political levels. This widening gap creates unequal opportunities, adds pressure to communities, and makes it harder for people to succeed in the job market. Education plays an important role in this situation. People with higher levels of education tend to earn more money than those with lower levels of education. But how straightforward is that relationship?

Several well-known sources have pointed out that income inequality is a major concern. The Organisation for Economic Co-operation and Development (OECD) has stated in multiple reports that the U.S. has one of the highest levels of income inequality among developed countries (Denk et al., 2013). They point out that this can slow down economic growth and limit social mobility. Horowitz et al. (2020) from the Pew Research Center has also found that income gaps keep growing and that education is a key factor—people without higher education are falling further behind in the labor market. Denk et al. (2013) and Horowitz et al. (2020) show that income inequality, and the role education plays in it, is a long-term issue that needs real attention.

Even tough this topic has been researched before, this analysis takes a closer look than many past studies. We focus on differences between U.S. states to see how the link between education and income changes from one region to another. By doing this, we can offer a more detailed and complete picture of income inequality. With this more in-depth analysis, we hope to find new insights that haven’t been explored in other research.

<https://www.oecd.org/en/publications/inequality-and-poverty-in-the-united-states_5k46957cwv8q-en.html>

# 2 Data Sourcing

## 2.1 Description and limitations

The dataset “Inequality and Growth in the United States: Evidence from a New State-Level Panel of Income Inequality Measure” of the researcher Frank has been used for our project. This dataset provides historical coverage (1990-2000) of the highest income group in the US. Another reason why this data set has been used is because it specifically shows the top income groups per state. This way, the differences per state can be easily compared to one another. The dataset shows multiple variables that can measure inequality. The data is not from a government website, but from a researcher himself. However, the researcher did use data from the World Inequality Database to calculate the different top income groups. The limitation of this dataset is that it only provides the top income percentages, so there is not a possibility to compare the bottom 10 percent with the top 10, which makes it harder to accurately measure inequality.

The second dataset is “The Educational attainment of persons 25 years old and over, by race/ethnicity and state: April 1990 and April 2000”. This dataset focuses on education level and has been retrieved from the National Center form Education Statistics (NCES). We chose this dataset because the level of education is a key indicator of inequality. The higher the education, the higher the income. Therefore big differences between education level will lead to a greater inequality. Furthermore, the dataset specifies the education level across all U.S. states. Which we can combine with the dataset of the top income percentages. Differences across racial and ethnic groups are also shown in the data, but will not be considered here, as they fall outside the scope of this analysis. The limitation of the dataset is that it includes data about only two points in time: April 1990 and April 2000. Unfortunately, there is no data available for the years in between.

## 2.2 Load in the data

To load the datasets, a link is provided to a Google Drive folder, from which a ZIP file containing the two datasets can be downloaded.

```{r loading data, include=FALSE}
zipfile <- "datasets_to_share.zip"

if (!file.exists(zipfile)) {
  download.file("https://drive.google.com/uc?export=download&id=1TYJDCbxCDWhbZwQ6kkY-RJXtwDgOMdX2",
                destfile = zipfile, mode = "wb")}

unzip(zipfile, exdir = "datasets_income_inequality")
```

```{r}
#Load the files
Distribution <- read_excel("datasets_income_inequality/Frank_WID_2020.xls", sheet = 3)
Education <- read_excel("datasets_income_inequality/tabn012.xls", col_types = c("text", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip","skip", "skip", "skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip"))

```

# 3 Quantifying

## 3.1 Data Cleaning

Only the years 1990 to 2000 are kept in the Distribution dataset. This is because the Education dataset includes data only for 1990 and 2000. To ensure the datasets can be merged and compared properly, just 1990, 2000, and the years in between are included.

In addition, two rows are removed, as they are not a U.S. state. One of the rows corresponds to the entire United States, while the other represents the District of Columbia.

```{r}
Distribution <- Distribution %>%
  filter(Year %in% 1990:2000, 
         State != c("United States", "District of Columbia"))
```

For the Education dataset, unnecessary rows are removed. These rows contain no data and consist only of blank spaces between entries. Also, two rows representing the entire United States and the District of Columbia are removed.

To simplify the datatset, columns are renamed to reflect the type of data they contain, and any dots following state names are removed.

For the final step of cleaning, before merging the two datasets, the Education dataset needs to be converted to a long format. Currently, it contains four separate variables for the years 1990 and 2000, but these need to be combined into a single column with the years listed vertically. This transformation allows the dataset to be merged properly with the Distribution dataset, which is already in long format.

```{r}
Education <- Education[ -c(1:13, 14, 20, 24, 26, 32, 38, 44, 50, 56, 62, 68, 75:79), ]
```

```{r}
names(Education)[1:5] <- c("State", "Highschool1990", "Highschool2000", "Bachelor1990", "Bachelor2000")
```

```{r}
Education <- Education %>% 
  mutate(State = State %>%
           str_replace_all("…+", "") %>%
           str_replace_all("\\.+", "") %>%
           str_trim()) # to delete all the dots after the state 
```

```{r}
df1990 <- data.frame(State = Education$State, Year = 1990, Highschool = Education$Highschool1990, Bachelor = Education$Bachelor1990, stringsAsFactors = FALSE)

df2000 <- data.frame(State = Education$State,Year = 2000,Highschool = Education$Highschool2000,Bachelor = Education$Bachelor2000,stringsAsFactors = FALSE)

Education <- rbind(df1990, df2000)

Education <- Education %>%
  mutate(across(3:4, ~ as.numeric(.)))

```

To merge the datasets, the full_join() function is applied. This function is appropriate because it performs an outer join, retaining all observations from both datasets based on the specific join keys (e.g., State and Year). Since the distribution dataset includes data for each year from 1990 to 2000, and the Education dataset only includes data for 1990 and 2000, full_join() ensures that no rows are lost during the merge. Missing values are introduced where data is not available in one of the datasets, allowing for a complete and alinged structure for subsequent analysis. After the merge, all numeric values are rounded to two decimals.

```{r}
Inequality <- Distribution %>%
  full_join(Education, by = c("State", "Year"))

Inequality <- Inequality %>%
  select(-st) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))
```

## 3.2 Describing the type of variables in the datasets

```{r}
head(Inequality)
```

This is the merged dataset, named "Inequality". The first variable indicates the year (i.e., from 1990 to 2000), and the second column identifies the state from which the data originates. Each state appears eleven times in the dataset - once for every year.

The next five variables represent the percentage of total income earned by specific income groups. For example, the first of these columns shows that the top 10 percent of earners in Alabama in 1990 earned 39.9% of the total income. The following four columns follow the same structure, with each representing increasingly smaller income groups.

The final two columns capture educational attainment levels, expressed in percentages. For instance, in 1990, 66.9% of the population in Alabama had a high school diploma or higher and 15.7% held a bachelor's degree or higher.

## 3.2 Creating new variables

To measure income inequality, a new variable is constructed, modeled after the Palma Ratio. This ratio is defined as "the ratio of the income share of the top 10 percent over that of the bottom 40 percent" (Basu & Stiglitz, 2016, p. xxvi), and is designed to capture inequality at the extremes of the income distribution. However, the Distribution dataset does not include the income share of the bottom 40 percent. Therefore, as a practical and intuitive alternative, inequality is measured by dividing the income share of the top 10 percent by that of the bottom 90 percent (i.e., 100% minus the top 10%).

Using this measure, inequality will be visualized over time by presenting the mean and median values of all U.S. states. This visualization provides a broad overview of the national trend in inequality across the country, showing whether inequality is generally increasing or decreasing over time.

Additionally, a second new variable groups states into four categories - Low, Medium Low, Medium High, and High - based on the percentage of the population with a bachelor's degree or higher, allowing inequality to be examined across sub-populations. This approach helps reveal whether states with a higher share of highly educated residents tend to have a higher or lower inequality.

Finally, inequality will be visualized across the 50 states through a third new variable that shows the absolute change in inequality between 1990 and 2000. Not all states experience inequality changes equally; some show significant increases while others remain stable. This variation can be attributed to factors such as local institutions, tax policies, or demographic shifts (Wei, 2015; Arcabic et al., 2021).

```{r}
Inequality <- Inequality %>%
  mutate(Top10_vs_bottom90 = round(Top10_adj/(100-Top10_adj), 2))
```

```{r}
Inequality <- Inequality %>%
  group_by(Year) %>%
  mutate(Quartile = ntile(Bachelor, 4), Bachelor_level = case_when(Quartile == 1 ~ "Low", Quartile == 2 ~ "Medium Low", Quartile == 3 ~ "Medium High",Quartile == 4 ~ "High"))

# Make sure Education is a factor with the right order
Inequality$Bachelor_level <- factor(
  Inequality$Bachelor_level,
  levels = c("Low", "Medium Low", "Medium High", "High"))
```

## 3.3 Temporal Variation

```{r echo=FALSE}
df_summary <- Inequality %>%
  group_by(Year) %>%
  summarise(median_ineq = median(Top10_vs_bottom90),
            mean_ineq = mean(Top10_vs_bottom90))

ggplot(df_summary, aes(x = Year)) +
  geom_line(aes(y = median_ineq), color = "blue", 
            linewidth = 1.2, 
            lineend = "round") +
  geom_line(aes(y = mean_ineq), 
            color = "red",
            linewidth = 1.2,
            lineend = "round") +
  labs(title = "Income Inequality per year", 
       subtitle = "Median (blue) vs Mean (red)",
       x = "\nYear", 
       y = "Top 10% / Bottom 90% ratio\n") +
  scale_x_continuous(limits = c(1990, 2000), breaks = 1990:2000) +
  scale_y_continuous(breaks = seq(0.5, 0.8, by = 0.02)) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))  

```

## 3.4 Sub-population Variation

```{r echo=FALSE}
library(ggplot2)


ggplot(data = Inequality[Inequality$Year == 2000, ],
       aes(x = Bachelor_level, y = Top10_vs_bottom90)) +
  geom_boxplot(fill = "lightblue", color = "black", width = 0.6, outlier.shape = 16, outlier.color = "gray40") +
  labs(
    x = "\nEducation Group (Based on % with Bachelor's Degree)",
    y = "Income Inequality\n(Top 10% vs Bottom 90%)\n",
    title = "Income Inequality by Education Level (Year 2000)"
  ) + 
   theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

```

## 3.5 Spatial Variation

```{r include=FALSE}
# Make sure these packages are installed and loaded
library(maps)


diff_data <- Inequality %>%
  filter(Year %in% c(1990, 2000)) %>%
  select(State, Year, Top10_vs_bottom90) %>%
  pivot_wider(names_from = Year, values_from = Top10_vs_bottom90) %>%
  mutate(
    # Extract only the real state name before any dots or extra characters
    state_lower = tolower(gsub("^([a-zA-Z ]+).*", "\\1", State)),
    state_lower = trimws(state_lower),
    diff = `2000` - `1990`
  ) %>%
  filter(state_lower != "district of columbia")

us_states <- map_data("state")
map_plot_data <- left_join(us_states, diff_data, by = c("region" = "state_lower"))
```

```{r include=FALSE}
# Check: should now match!
print(setdiff(sort(unique(us_states$region)), sort(unique(diff_data$state_lower))))
print(setdiff(sort(unique(diff_data$state_lower)), sort(unique(us_states$region))))
```

```{r echo=FALSE}

ggplot(map_plot_data, aes(long, lat, group = group, fill = diff)) +
  geom_polygon(color = "black") +
  coord_fixed(1.3) +
  scale_fill_gradient2(
    low = "blue", mid = "ivory", high = "red", midpoint = 0,
    name = "Difference in \nTop 10% vs Bottom 90% ratio\n(1990-2000)",
    limits = c(-0.2, 0.8)
  ) +
  labs(
    title = "Change in Income Inequality by State",
    subtitle = "(Top 10% vs Bottom 90%, 1990-2000)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.text.align = 0.6
  )
```

## 3.5 Event Analysis

```{r echo=FALSE}
df_summary <- Inequality %>%
  group_by(Year) %>%
  summarise(median_ineq = median(Top10_vs_bottom90),
            mean_ineq = mean(Top10_vs_bottom90))

ggplot(df_summary, aes(x = Year)) +
  geom_line(aes(y = median_ineq), color = "blue", 
            linewidth = 1.2, 
            lineend = "round") +
  geom_line(aes(y = mean_ineq), 
            color = "red",
            linewidth = 1.2,
            lineend = "round") +
  labs(title = "Income Inequality per year", 
       subtitle = "Median (blue) vs Mean (red)",
       x = "\nYear", 
       y = "Top 10% / Bottom 90% ratio\n") +
  scale_x_continuous(limits = c(1990, 2000), breaks = 1990:2000) +
  scale_y_continuous(breaks = seq(0.5, 0.8, by = 0.02)) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)) +
  geom_vline(xintercept = 1994, size = 1.2) +
  annotate("text", x = 1995, y = 0.77, size = 4, label = "Introduction\nof NAFTA")

```

The temporal visualization shows that inequality rose significantly after 1994, which may be linked to the implementation of the North American Free Trade Agreement (NAFTA). NAFTA, established between the U.S., MExico and Canada came into effect in January 1994 with the objective of creating a trilateral trade bloc to increase economic integration, boost trade, and stimulate growth by removing trade barriers and import tariffs. As a result, exports and imports became cheaper, and a significant share of U.S. manufacturing was relocated to Mexico due to lower labor costs (SOURCE).

NAFTA established the world’s largest free trade zone, but not everyone benefited equally. Higher-income groups in the U.S. gained the most, as large corporations experienced increased profits that mainly flowed to shareholders and high-income individuals (This trend explains the rise in the mean income, while the median grew more slowly)(SOURCE).

While the NAFTA boosted overall GDP, it disproportionately affected lower-income groups. The United States lost numerous manufacturing jobs (Estimated at 682,900 by 2010) as companies relocated production to Mexico to cut labor costs (Scott, 2011). This acceleration of outsourcing significantly impacted American manufacturing and low-skilled workers. Consequently, some workers saw their income stagnate or decline, whereas individuals in management and high-skill sectors experienced income growth (SOURCE). As a result, income inequality widened, further expanding the gap between the bottom 90 percent and the top earners.

However, it is important to recognize that the rise in income inequality cannot be attributed solely to NAFTA. Factors such as technological advancements, shifts in labor market policies and broader globalization have also played significant roles in driving this trend.

# 4 Discussion

## 4.1 Discussion of the findings

This project has shown that inequality in the US is structural growing and a more urgent social issue. The income share ratio that we constructed shows that there is an increase in inequality between 1990 and 2000. This increase is visible in almost every US state. The largest increases occurred in Massachusetts, California, Arizona, and Nebraska. The changes in inequality may be linked to regional economic factors in specific areas. For example California and Massachusetts had a rapid growth in tech and finance industries during the 1990s, where highly skilled workers (often in the top 10%) were paid significantly more than lower income groups.

The analysis that can be made from our dataset is that the income inequality started rising more starting from the year 1994. This year (NAFTA) was implemented. The goal of NAFTA was to stimulate trade and economic growth, but this agreement also led to outsourcing to lower-wage countries. Because of this outsourcing, many low-skilled manufacturing job in the US were lost. While the lower class was struggling, high income groups were benefiting from rising corporate profits and stock value. Which is why the inequality started to rise significantly more.

The sub-population analysis, based on the share of people with a bachelor’s degree or higher, showed that states who have a higher education level, often tend to have greater inequality. The connection between this relation is that highly educated workers have more access to better-paying jobs, which further widens the income gap.

Our analysis shows that inequality is widespread but most prominent in states with higher education levels and following policy shifts such as NAFTA.

# 5 Reproducibility

## 5.1 Github Repository Link

## 5.2 Reference List

Arčabić, V., Kim, K. T., You, Y., & Lee, J. (2021). Century-long dynamics and convergence of income inequality among the US states. Economic Modelling, 101, 105526. <https://doi.org/10.1016/j.econmod.2021.105526>

Denk, O., Hagemann, R. P., Lenain, P., Somma, V. (2013). Inequality and Poverty in the United States: Public Policies for Inclusive Growth. In *OECD Economics Department Working Papers* (No. 1052). <https://dx.doi.org/10.1787/5k46957cwv8q-en>

Horowitz, J., Igielnik, R., Kochhar, R., & Pew Research Center. (2020b). *Most Americans Say There Is Too Much Economic Inequality in the U.S., but Fewer Than Half Call It a Top Priority*. <https://www.pewsocialtrends.org/wp-content/uploads/sites/3/2020/01/PSDT_01.09.20_economic-inequailty_FULL.pdf>

Basu, K., & Stiglitz, J. E. (Eds.). (2016). Inequality and Growth: Patterns and Policy Volume II: Regions and Regularities. Palgrave Macmillan. <https://doi.org/10.1057/9781137554598>

Wei, Y. D. (2015). Spatiality of regional inequality. Applied Geography, 61, 1–10. https://dx.doi.org/10.1016/j.apgeog.2015.03.013

Scott, R. E. (2011). Heading South: U.S.-Mexico trade and job displacement after NAFTA (Briefing Paper #308). Economic Policy Institute.
