
---
title: "The Effect of Education Level on Income Inequality"
authors: "Names + Studentnumbers"
date: "`r Sys.Date()`"
output: pdf_document
---

#Set up your environment
```{r}
setwd("~/")
```

```{r packages_install, include = FALSE}
install.packages("tidyverse")
install.packages("renv")
install.packages("dplyr")
```

```{r}
library(tidyverse)
library(renv)
library(dplyr)
```

```{r}
packageVersion("tidyverse")
packageVersion("renv")
packageVersion("dplyr")
```

#Title Page
Carlijn Calori, Leah Delikát, Fadhil Dhafir, Sten Groen, Julia Koeleman, Anne Schrama, Marie-Louise Stevens, Sophia Zentgraf
Tutorial Group 1
J.F. Fitzgerald

#Part 1 - Identification of the Social Problem

##1.1 Describe the Social Problem
Income inequality in the USA is a social problem because it limits equal opportunities, increases social tensions, and harms health and democracy.The Pew Research Center warns that the middle class is shrinking rapidly, while the rich grow richer.In addition, The OECD ranks the U.S. among the most unequal developed nations. These two studies emphasize the importance of this problem ABCDEF 


https://www.pewresearch.org/social-trends/2020/01/09/trends-in-income-and-wealth-inequality/?
https://www.oecd.org/en/publications/inequality-and-poverty-in-the-united-states_5k46957cwv8q-en.html

#Part 2 - Data Sourcing

##2.1 Load in Data
URL VAN DATA SETS:

```{r}


```

```{r loading data}


```


##2.2 Provide a short summary of the datasets

```{r}
head ()

```

##2.3 
``` {r}

``` 
