
---
title: "The Effect of Education Level on Income Inequality"
authors: "Names + Studentnumbers"
date: "`r Sys.Date()`"
output: pdf_document
---

#Set up your environment
```{r}
setwd("~/")
```

```{r packages_install, include = FALSE}
install.packages("tidyverse")
install.packages("renv")
install.packages("dplyr")
```

```{r}
library(tidyverse)
library(renv)
library(dplyr)
```

```{r}
packageVersion("tidyverse")
packageVersion("renv")
packageVersion("dplyr")
```

#Title Page
Carlijn Calori, Leah Delikát, Fadhil Dhafir, Sten Groen, Julia Koeleman, Anne Schrama, Marie-Louise Stevens, Sophia Zentgraf
Tutorial Group 1
J.F. Fitzgerald

#Part 1 - Identification of the Social Problem

##1.1 Describe the Social Problem
Income inequality in the USA is a social problem because it limits equal opportunities, increases social tensions, and harms health and democracy.The Pew Research Center warns that the middle class is shrinking rapidly, while the rich grow richer.In addition, The OECD ranks the U.S. among the most unequal developed nations. These two studies emphasize the importance of this problem ABCDEF 


https://www.pewresearch.org/social-trends/2020/01/09/trends-in-income-and-wealth-inequality/?
https://www.oecd.org/en/publications/inequality-and-poverty-in-the-united-states_5k46957cwv8q-en.html

#Part 2 - Data Sourcing

##2.1 Load in Data
URL VAN DATA SETS:


```{r loading data}
library(readxl)
Distribution <- read_excel("data/Frank_WID_2020.xls", sheet = 3)
Tresholds <- read_excel("data/Frank_WID_2020.xls", sheet = 4)
Education <- read_excel("data/tabn012.xls", col_types = c("text", "skip", "text", 
"skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", 
"skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", "skip", "skip", 
"skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", "skip", "skip", 
"skip", "skip", "skip", "skip", "skip", "text", "skip", "skip", "skip", "skip", 
"skip", "skip", "skip", "skip", "skip", "skip"))

```

Clean the datasets
```{r}

Distribution <- Distribution %>%
  filter(Year %in% c(1990, 2000), 
         State != c("United States", "District of Columbia")) %>%
  mutate(State_ID = 1:100)

Tresholds <- Tresholds %>%
  filter(Year %in% c(1990, 2000), 
         State != c("United States", "District of Columbia")) %>%
  mutate(State_ID = 1:100)

Education <- Education[ -c(1:13, 14, 20, 24, 26, 32, 38, 44, 50, 56, 62, 68, 75:79), ]

names(Education)[1:5] <- c("State", "Highschool1990", "Highschool2000", "Bachelor1990", "Bachelor2000")
```

Reshape the Education dataset to the same form as the others so you have a variable Year, and for every state a row in 1990 and 2000
```{r}
df1990 <- data.frame(State = Education$State, Year = 1990, Highschool = Education$Highschool1990, Bachelor = Education$Bachelor1990, stringsAsFactors = FALSE)

df2000 <- data.frame(State = Education$State,Year = 2000,Highschool = Education$Highschool2000,Bachelor = Education$Bachelor2000,stringsAsFactors = FALSE)

Education <- rbind(df1990, df2000)

Education <- Education1 %>%
  arrange(Year, State) %>%
  mutate(State_ID = 1:100) %>%
  mutate(across(3:4, ~ as.numeric(.)))

#Or more efficient thanks to chatgpt:

Education <- Education %>%
  mutate(across(2:5, ~ as.numeric(.))) %>%
  pivot_longer(
    cols = -State,
    names_to = c(".value", "Year"),
    names_pattern = "(Highschool|Bachelor)(\\d+)") %>%
  arrange(Year) %>%
  mutate(
    State_ID = 1:100, 
    Year = as.numeric(Year))
```

To merge the datasets:
```{r}
Allmerged <- Distribution %>%
  inner_join(Tresholds, by = c("State_ID", "Year")) %>%
  inner_join(Education, by = c("State_ID", "Year"))

Allmerged <- Allmerged %>%
  select(-st.x, -st.y, -State.x, -State.y, -State) %>%
  relocate(State_ID, .after = Year) %>%
  mutate(across(3:16, ~ round(.x, 2)))

#North Carolina has state_ID 33 for 1990, and 33 + 50 (number of states) = 83 for 2000 and Nevada for example 28 for 1990 and 28 + 50 = 78 for 2000. Idk if this form of identification makes sense, but i think so

```


##2.2 Provide a short summary of the datasets

```{r}
head ()

```

##2.3 
``` {r}

``` 
